---
title: "SNEMA Winter Flounder EDM"
author: "tara"
date: "9/1/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning=FALSE, message=FALSE}
#install.packages("devtools") #if required
#devtools::install_github("tanyalrogers/GPEDM") #update frequently
library(dplyr)
library(tidyverse)
library(purrr)
library(ggplot2)
library(rEDM)
library(GPEDM)
library(Metrics)
library(kableExtra)
library(corrplot)
source("hierEDM_helper_functions.R")
```


The idea here is to look at dynamic correlation of winter flounder indices to see connectivity. Maybe we can map it as a function of distance? 
Another idea would be to try to revisit the predator/prey and the environmental correlates. 



**Load Data**  
Are these "mean numbers per tow", such as the nmfs bts, expanded with an age-length key?   
Do they take changes in gear/ship etc. into account?  
Not always clear whether 7 is a plus group. Perhaps cut to age 6?
**Why are we predominantly focused on spring surveys, when it's most likely the fish are up in the estuaries at that time???**  
We should get some fall survey data.
```{r}
dmf_spring <- read.csv("data_WF_SNEMA/DMFSPRING_WFSNEMA.csv", header = TRUE)##MADMF Spring BTS mean numbers per tow, 7 is a plus group
ct_spring <- read.csv("data_WF_SNEMA/CTSpring_WFSNEMA.csv", header = TRUE) #CTDEEP spring BTS mean numbers per tow, 7 is a plus group
nmfs_fall <- read.csv("data_WF_SNEMA/fallBTS_WFSNEMA.csv", header = TRUE) #NMFS fall BTS mean numbers per tow, 7 is a plus group
nmfs_spring <- read.csv("data_WF_SNEMA/springBTS_WFSNEMA.csv", header = TRUE)#NMFS Spring BTS mean numbers per tow, 7 is a plus group
nj_river <- read.csv("data_WF_SNEMA/NJRiver_WFSNEMA.csv", header = TRUE)#mean numbers per tow, 7 is a plus group
nj_trawl <- read.csv("data_WF_SNEMA/NJTrawl_SNEMAWF.csv", header = TRUE)#mean numbers per tow, 7 is a plus group
ri_spring <- read.csv("data_WF_SNEMA/RI_Spring_SNEMAWF.csv", header = TRUE) # RI DFW spring mean numbers at age. 
recruits <-read.csv("data_WF_SNEMA/SNEMAWF_recruitment.csv", header = TRUE) #mean numbers per tow recruitmet (age 0) indices
uri_trawl <-read.csv("data_WF_SNEMA/URIGSO_WF.csv", header = TRUE) #mean numbers per tow, 7 is a plus group
nemap_spring <-read.csv("data_WF_SNEMA/wfsurvey_NEMAP_SPRING.csv", header = TRUE) #nemap, 7 is a plus group
nmfs_winter <-read.csv("data_WF_SNEMA/winter_bts_SNEMAWF.csv", header = TRUE) #NMFS winter BTS mean numbers per tow, 5 is a plus group
comm_landings <-read.csv("data_WF_SNEMA/commercial_landings_WFSNEMA.csv", header = TRUE) #commercial landings (no discards, kg, mt and lbs)
total_catch <-read.csv("data_WF_SNEMA/total_catch_SNEMAWF.csv", header = TRUE) #total catch 

#discards, MRIP, MRIP CAA< WAA, SSB_WAA, January 1 WAA,  data also available. 

```

Clean data  
```{r}
#format the survey datasets identically - 
#####does not include recruits, nmfs_winter, comm landings, total catch
df_list <- list(dmf_spring,ct_spring,nmfs_fall,nmfs_spring,nj_river,ri_spring,uri_trawl, nemap_spring)
names(df_list) <-c("dmf_spring","ct_spring","nmfs_fall","nmfs_spring","nj_river","ri_spring","uri_trawl", "nemap_spring")
TA_list <-list()
df_list2 <-list()
df_list_np <-list()
for (i in 1:length(df_list)){
  TA_list[[i]] <-dplyr::select(df_list[[i]], Year, Total)%>%mutate(survey=names(df_list)[i])
  df_list2[[i]] <-dplyr::select(df_list[[i]], 1:9)%>%mutate(survey=names(df_list)[i])
  names(df_list2[[i]]) <-c("Year","Age_1","Age_2","Age_3","Age_4","Age_5","Age_6","Age_7plus","Total")
  df_list_np[[i]] <-dplyr::select(df_list2[[i]], 1:7)%>%mutate(survey=names(df_list)[i]) #version without total and without plus group. 
  names(df_list_np[[i]]) <-c("Year","Age_1","Age_2","Age_3","Age_4","Age_5","Age_6","survey")
}

#create a dataset with total abundance for all the locations
total_abundance <-bind_rows(TA_list)

#create a long form dataset with all years and surveys combined. 
all_surveys <-bind_rows(df_list_np)
allsurv_long <-pivot_longer(all_surveys, cols=2:7, names_to = "age_class",values_to = "abundance")

```

Plot data
```{r}
allsurv_long%>%
  ggplot(aes(Year,abundance, color=survey))+
  geom_line()+
  facet_wrap(~age_class, scales="free")+theme_classic()

allsurv_long%>%
  ggplot(aes(Year,abundance, color=survey))+
  geom_line()+
  facet_grid(survey~age_class, scales="free")+theme_bw()

```  

**Consider removing NEMAP_spring and NJ river and trimming data to 1985-2021**   
because there's not very many years of data there. 
```{r}
##min and max year for each survey
allsurv_long%>%group_by(survey)%>%summarize(maxyear=max(Year),minyear=min(Year))

allsurv_long <-filter(allsurv_long, survey != "nemap_spring" & survey != "nj_river")%>%filter(Year >=1985)
total_abundance <-filter(total_abundance, survey != "nemap_spring" & survey != "nj_river")%>%filter(Year >=1985)

```

**DECIDE WHETHER TO LOG THE DATA**  
skip this chunk if you are not logging. I think we will NOT log the data because of the +1 issue. 
```{r, eval=FALSE}
#allsurv_log <-mutate(allsurv_long, abundance=log(abundance+1))
#TA_log <-mutate(total_abundance, N_total=log(Total+1))
```

**CORRELATION**  
lagged correlation of age classes within each survey
surveys within each age class  
total abundance by survey  
```{r}
#total abundance by survey  
ta_wide <-pivot_wider(total_abundance,id_cols=Year, names_from =survey, values_from = Total)
M <-cor(ta_wide[,-1], use="pairwise.complete.obs")
corrplot(M, type="upper", method="color", title="total abundance", mar=c(0,0,2,0), addCoef.col = 'black', diag=F,tl.col = "black")

#surveys within each age class
age_list <- colnames(all_surveys)[2:7]
for (i in 1:length(age_list)){
  df <-all_surveys[,c(1,i,8)]
  df_wide <-pivot_wider(df, id_cols=Year, names_from = survey, values_from=names(df)[2])
  M <-cor(df_wide[,-1], use="pairwise.complete.obs")
  corrplot(M, type="upper", method="color", title=age_list[[i]], mar=c(0,0,2,0), addCoef.col = 'black', diag=F,tl.col = "black")
}

#lagged age classes within each survey
surv_list <-unique(all_surveys$survey)
surv_list<-surv_list[!surv_list %in% c("nj_river","nemap_spring")]
for (i in 1:length(surv_list)){
  df <- filter(all_surveys, survey==surv_list[[i]])
  dfblock <-df[,1:7]
  dfblock <-make_block(dfblock, max_lag = 6, tau=1)
  dfblock <-dplyr::select(dfblock,"Age_1(t+0)", "Age_2(t+1)","Age_3(t+2)","Age_4(t+3)","Age_5(t+4)","Age_6(t+5)")
  M <-cor(dfblock, use="pairwise.complete.obs")
  corrplot(M, type="upper", method="color", title=surv_list[[i]], mar=c(0,0,2,0), addCoef.col = 'black', diag=F,tl.col = "black")
}

```

**Total abundance fits across a grid of E and Tau**  
```{r, cache=TRUE}
Ees <-seq(1,5,1)
taus <-seq(1,3,1)
var_pairs = expand.grid(Ees, taus) # Combinations of vars, 2 at a time
ETdf <-matrix(nrow=dim(var_pairs)[1],ncol=4)
ETdf[,1]<-var_pairs[,1]
ETdf[,2]<-var_pairs[,2]
r2matrixTA = array(NA, dim = c(length(Ees), length(taus)), dimnames = list(Ees,taus)) 
rmsematrixTA = array(NA, dim = c(length(Ees), length(taus)), dimnames = list(Ees,taus)) 
for (i in 1:nrow(var_pairs)) {
 try({
  fitTA <-fitGP(data = total_abundance, y = "Total", pop="survey",scaling = "local", E=var_pairs[i,1], tau=var_pairs[i,2], predictmethod = "loo")
  fitTA_r2 <-fitTA$outsampfitstats[[1]]
  fitTA_rmse <-fitTA$outsampfitstats[[2]]
  r2matrixTA[var_pairs[i,1], var_pairs[i,2]] = fitTA_r2
  ETdf[i,3] <-fitTA_r2
  ETdf[i,4] <-fitTA_rmse
  rmsematrixTA[var_pairs[i,1], var_pairs[i,2]] = fitTA_rmse
 },silent=T)
}

#r2matrixTA
#rmsematrixTA

ETdf <-as.data.frame(ETdf)
names(ETdf)<-c("E","tau","OOS_R2","OOS_RMSE")

### Plot
#R- squared. 
ETdf %>%
 #filter(E != 10)%>%
 ggplot(aes(E,tau,label=round(OOS_R2,2)))+
 geom_tile(aes(fill=OOS_R2), show.legend = TRUE)+
 scale_fill_gradient(low="white",high="blue")+
 #scale_x_discrete(limits=seq(1,5,1))+
 geom_text()+
 xlab("embedding dimension")+ylab("tau")+
 guides(fill=guide_colorbar(title="r-squared"))+
 theme_bw()+
 theme(axis.text.x = element_text(size=12),
       axis.text.y = element_text(size=12),
       panel.background = element_rect(fill = "transparent", colour = "black"),
       strip.background =element_rect(fill="white"),
       panel.grid.major = element_blank(), 
       panel.grid.minor = element_blank())

#RMSE. 
ETdf %>%
 #filter(E != 10)%>%
 ggplot(aes(E,tau,label=round(OOS_RMSE,2)))+
 geom_tile(aes(fill=OOS_RMSE), show.legend = TRUE)+
 scale_fill_gradient(low="blue",high="white")+
 #scale_x_discrete(limits=seq(1,5,1))+
 geom_text()+
 xlab("embedding dimension")+ylab("tau")+
 guides(fill=guide_colorbar(title="RMSE"))+
 theme_bw()+
 theme(axis.text.x = element_text(size=12),
       axis.text.y = element_text(size=12),
       panel.background = element_rect(fill = "transparent", colour = "black"),
       strip.background =element_rect(fill="white"),
       panel.grid.major = element_blank(), 
       panel.grid.minor = element_blank())

``` 
R-squared indicates best E=2 and best tau=1  
RMSE indicates best E=5 and best tau=3

**fit models and extract fitstats - all of the data**  
```{r}
maxE=5

#total abundance combined across surveys - an overall index
ta_all <-total_abundance %>% group_by(Year)%>%summarize(TA=sum(Total))%>%as.data.frame()
TA_all <- fitGP(data=ta_all, y="TA", E=maxE, tau=1, predictmethod = "loo")

#total abundance, survey is the structure
TA_survey <-fitGP(data=total_abundance, y="Total", pop="survey", scaling="local", E=maxE, tau=1, predictmethod = "loo")

##### extract fitstats ####
#################################
outsamp <-bind_rows(TA_all$outsampfitstats,TA_survey$outsampfitstats)
names(outsamp) <-c("OOS_R2","OOS_rmse")
insamp <-bind_rows(TA_all$insampfitstats,TA_survey$insampfitstats)
rhos <-c(tail(TA_all$pars,1), tail(TA_survey$pars,1))
fitstats <-bind_cols(outsamp,insamp,rhos)%>%as.data.frame()
rownames(fitstats) <-c("total_abundance_combined","TA_by_survey")
colnames(fitstats)[11]<-"rho"
fitstats <-rownames_to_column(fitstats, var="model")%>%mutate(`age class`="all")%>% mutate(across(OOS_R2:df, as.numeric))
#indv surveys
TA_surveys <-full_join(TA_survey$outsampresults, TA_survey$insampresults, by=c("timestep","pop","obs"))%>%dplyr::rename(survey=pop)%>%mutate(year=timestep+1984)

#### Visualize fits over data ####
###### PLOT ######
TA_surveys%>%
 #filter(age !="total abundance")%>% #filter out N-total
 ggplot() +
 facet_wrap(survey~., scales = "free") +
 geom_line(aes(x=year,y=predmean.x)) + #out of sample
 #geom_line(aes(x=timestep,y=predmean.y),lty="dashed") + #insample
 geom_ribbon(aes(x=year,y=predmean.x,ymin=predmean.x-predfsd.x,ymax=predmean.x+predfsd.x), alpha=0.4,fill="#03C0C1") + #out of sample
 #geom_ribbon(aes(x=timestep,y=predmean.y, ymin=predmean.y-predfsd.y,ymax=predmean.y+predfsd.y), alpha=0.4,fill="#007F80") + #in sample
 geom_point(aes(x=year, y=obs)) +
 theme_classic()

```

**pairwise dynamic correlation between surveys**  
```{r}
maxE=5
#total abundance by survey  
vars = colnames(ta_wide[2:7])
var_pairs = combn(vars, 2) # Combinations of vars, 2 at a time
rho_matrixTA = array(NA, dim = c(length(vars), length(vars)), dimnames = list(vars,vars)) 
for (i in 1:ncol(var_pairs)) {
 df = filter(total_abundance, survey %in% c(var_pairs[1,i], var_pairs[2,i]))
 fitTA <-fitGP(data = df, y = "Total", pop="survey",scaling = "local", E=maxE, tau=1, predictmethod = "loo")
 fitTA_rho <-tail(fitTA$pars,1)
 rho_matrixTA[var_pairs[1,i], var_pairs[2,i]] = fitTA_rho
}
rho_matrixTA
round(rho_matrixTA,4) %>% 
 kbl()%>%kable_classic(full_width = F, html_font = "Cambria")

corrplot(rho_matrixTA, method="color", type="upper", diag=FALSE, is.corr = TRUE,mar=c(0,0,2,0),addCoef.col = 'black',tl.col = "black")

```

Pairwise dynamic correlation: surveys within each age class - **Can't get this to work for now**. 
```{r, eval=F}
#surveys within each age class
all_surveys2 <-filter(all_surveys, survey != "nemap_spring" & survey != "nj_river")%>%
   filter(Year > 1984 & Year < 2022)
age_list <- colnames(all_surveys2)[1:ncol(all_surveys2)-1][-1]
rm_list <-list()
for (i in 1:length(age_list)){
  df <-all_surveys2[,c(1,i,8)]
  df_wide <-pivot_wider(df, id_cols=Year, names_from = survey, values_from=names(df)[2])%>%
    mutate(age=colnames(all_surveys2)[[i]])
  
vars = colnames(df_wide)[2:ncol(df_wide)-1][-1]
var_pairs = combn(vars, 2) # Combinations of vars, 2 at a time
rho_matrixTA = array(NA, dim = c(length(vars), length(vars)), dimnames = list(vars,vars)) 

for (j in 1:ncol(var_pairs)) {
 df2 = filter(df, survey %in% c(var_pairs[1,j], var_pairs[2,j]))
colnames(df2)[2] <-"Total"
 fitTA <-fitGP(data = df2, y = "Total", pop="survey",scaling = "local", E=maxE, tau=1, predictmethod = "loo")
 fitTA_rho <-tail(fitTA$pars,1)
 rho_matrixTA[var_pairs[1,j], var_pairs[2,j]] = fitTA_rho
 
}
rm_list[[i]]<-rho_matrixTA
}
```



**fit models - predict last five years**

**fit models - environmental data**

