---
title: "WFpreds2"
author: "tara"
date: "9/26/2023"
output: html_document
---

This is a version of WFPreds.Rmd that we are using specifically for the SMAST talk on 10/4/23 - Sticking to striped bass and summer flounder. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(kableExtra)
library(forcats)
library(Metrics)
library(tidyverse)
library(purrr)
library(rEDM)
library(corrplot)
#setwd("C:/Users/tdolan/Documents/R-Github/age_structureEDM")
```

**CTDEEP Striped bass data - I think this is the spring index? we need to find out. We would be interested in separately looking at fall and spring.** 
```{r}
#using CTSB indices for now
CTSB <- read.csv("CTtrawlabundanceage19872017 copy.csv", header = TRUE)

#we are getting rid of the 15+ category
CTSB <-CTSB[,2:16]
pop24 <-pivot_longer(CTSB, 2:15, names_to = "age_class")%>%mutate(age_class=as.factor(age_class),lNvalue=log(value))%>%as.data.frame()%>%mutate(age_class = fct_relevel(age_class, c("X1","X2","X3","X4","X5","X6","X7","X8","X9","X10","X11","X12","X13","X14")))

#create an abundance index from all age classes
popAll <-mutate(CTSB, N_total=X1+X2+X3+X4+X5+X5+X6+X7+X8+X9+X10+X11+X12+X13+X14)
popAll <-mutate(popAll, Year=as.integer(Year))

#make a vertical version of popALL which you will use later. 
names(popAll)<-c("time_step","1","2","3","4","5","6","7","8","9","10","11","12","13","14","N_total")
popvert <-as.data.frame(pivot_longer(popAll, 2:15, names_to = "age",values_to="abundance"))
popvert <-popvert %>% mutate(time_step=as.numeric(time_step)) %>%
                             group_by(time_step) %>% arrange(age) %>%as.data.frame()

#Make a scaled dataset
means <-colMeans(popAll[sapply(popAll, is.numeric)])
stdevs <-popAll %>% summarise_if(is.numeric, sd)
savescale <-as.data.frame(cbind(means,t(stdevs)))
names(savescale)<-c("means","sd")
savescale <-savescale[2:15,]
popscale <-transmute_at(popAll,2:15, scale) 
popscale <-rownames_to_column(popscale, var="time_step")

#vertical scaled dataset
popscv <-as.data.frame(pivot_longer(popscale, 2:15, names_to = "age",values_to="abundance"))
popscv <-popscv %>% mutate(time_step=as.numeric(time_step)) %>%
                             group_by(time_step) %>% arrange(age)%>%as.data.frame()
```

**CTDEEP Summer Flounder Data - Spring because correlated to early season.** 
```{r}
#1984-2012
CTSF_spring <-read.csv("SF_spring_CTDEEP.csv", header = TRUE)

#format the data
#we are getting rid of the 15+ category
CTSF <-CTSF_spring[,c(1,3:8)]
popCTSF <-pivot_longer(CTSF, 2:7, names_to = "age_class")%>%mutate(age_class=as.factor(age_class),lNvalue=log(value))%>%as.data.frame()%>%mutate(age_class = fct_relevel(age_class, c("X1","X2","X3","X4","X5","X6")))

#create an abundance index from all age classes
popAllCTSF <-mutate(CTSF, N_total=X1+X2+X3+X4+X5+X5+X6)
popAllCTSF <-mutate(popAllCTSF, Year=as.integer(Year))

#make a vertical version of popALL which you will use later. 
names(popAllCTSF)<-c("time_step","1","2","3","4","5","6","N_total")
popvert_CTSF <-as.data.frame(pivot_longer(popAllCTSF, 2:8, names_to = "age",values_to="abundance"))
popvert_CTSF <-popvert %>% mutate(time_step=as.numeric(time_step)) %>%
                             group_by(time_step) %>% arrange(age) %>%as.data.frame()

#Make a scaled dataset
means <-colMeans(popAllCTSF[sapply(popAllCTSF, is.numeric)])
stdevs <-popAllCTSF %>% summarise_if(is.numeric, sd)
savescale <-as.data.frame(cbind(means,t(stdevs)))
names(savescale)<-c("means","sd")
savescale <-savescale[2:8,]
popscaleCTSF <-transmute_at(popAllCTSF,2:8, scale) 
popscaleCTSF <-rownames_to_column(popscaleCTSF, var="time_step")

#vertical scaled dataset
popscv_CTSF <-as.data.frame(pivot_longer(popscaleCTSF, 2:8, names_to = "age",values_to="abundance"))
popscv_CTSF <-popscv_CTSF %>% mutate(time_step=as.numeric(time_step)) %>%
                             group_by(time_step) %>% arrange(age)%>%as.data.frame()
```

**CTDEEP winter flounder YOY data**  
```{r}
#spring survey 1988-2016
WFCTYOY <-read.csv("WF_CTYOY_matt.csv", header = TRUE)
#age 0 shifted forward one year making it an age 1 index (according to Matt)
WFCTYOY <-WFCTYOY[8:36,1:2]
WFCTYOY <-mutate(WFCTYOY, age0 = lead(Total,1)) %>%select(-Total)#CAREFUL HERE 
```

**CTDEEP WF data**  
```{r}
WFCT <-read.csv("WF_CTDEEP_SPRING.csv", header = TRUE)
```

**Within CT, look at spring SF & SB to WF yoy** 
```{r}
#combine the datasets
#not sure if we should use scaled datasets or not.
popvert<-mutate(popvert, species="SB")
popvert_CTSF <- mutate(popvert_CTSF, species="SF")
ct_spring <- WFCTYOY %>% dplyr::rename(time_step=Year, abundance=age0)%>%
  mutate(age="0",species="WF")%>% bind_rows(popvert)%>%bind_rows(popvert_CTSF)%>%
  filter(time_step > 1987 & time_step < 2016)

#Try first sum of all adult SB and SF compared to WFYOY
WFCTYOY2 <-WFCTYOY %>% dplyr::rename(time_step=Year, WF=age0)
ct_yoy_allAd <-pivot_wider(ct_spring, names_from = species, values_from = N_total)%>%
  filter(age !="0")%>% select(-WF)%>%left_join(WFCTYOY2)%>%group_by(time_step)%>%
  summarize(SB=mean(SB),SF=mean(SF),WF=mean(WF))


#find embedding dimension
rho_E <- EmbedDimension(dataFrame = ct_yoy_allAd, columns = c("SB","SF","WF"), target = "WF",lib = "1 28", pred = "1 28", showPlot = TRUE) 

```
We will use E=3

```{r}
vars = colnames(ct_yoy_allAd[2:4])
var_pairs = combn(vars, 2) # Combinations of vars, 2 at a time

libSize = paste(10, 100, 5, collapse = " ") #have a standard library size
ccm_matrix = array(NA, dim = c(length(vars), length(vars)), dimnames = list(vars,vars)) 

for (i in 1:ncol(var_pairs)) {
ccm_out = CCM(dataFrame = ct_yoy_allAd, columns = var_pairs[1, i], target = var_pairs[2,i], 
              libSizes = libSize, tau= 1, E = 3,showPlot = TRUE, sample=100,replacement=TRUE) 
outVars = names(ccm_out)
var_out = unlist(strsplit(outVars[2], ":")) 
ccm_matrix[var_out[2], var_out[1]] = ccm_out[nrow(ccm_out), 2]
var_out = unlist(strsplit(outVars[3], ":"))
ccm_matrix[var_out[2], var_out[1]] = ccm_out[nrow(ccm_out), 3]
}

ccm_matrix

```

Nothing promising here. 

**every age of winter flounder against every age of SB in CT Spring**  
I am not sure that spring is the best index for the sound because they will be up in the estuaries not out in the sound. 
```{r}



```

**Peconics datasets**  
```{r}


```




**Evaluate the correlation dredging study**  
I am not sure this will work because the data appear to be scaled and we don't have the ability to unscale. 
```{r}
north_bays <-read.csv("northbays.csv", header = TRUE)%>%dplyr::rename(Survival=Survial, SF=SFRED, SST=SSTred)

#find embedding dimension - do YOYEarly and YOYLate separately. 

#YOYEarly
rho_E <- EmbedDimension(dataFrame = north_bays, columns = c("YOYEarly","Survival","TaveE","TmaxE","TminE",
            "TaveL","TmaxL","TminL","SST","SF","SB","BF","BC"), 
            target = "YOYEarly",lib = "1 29", pred = "1 29", showPlot = TRUE) 

#YOYLate
rho_E <- EmbedDimension(dataFrame = north_bays, columns = c("YOYLATE","Survival","TaveE","TmaxE","TminE",
            "TaveL","TmaxL","TminL","SST","SF","SB","BF","BC"), 
            target = "YOYEarly",lib = "1 29", pred = "1 29", showPlot = TRUE) 

#luckily E=3 works for both. - This also makes intuitive sense because age at maturity. 

#make a regular correlation plot
M <-cor(north_bays[,-1], use="pairwise.complete.obs")
corrplot(M, type="upper", method="color", title="", mar=c(0,0,2,0), 
         #addCoef.col = 'black', 
         diag=F,tl.col = "black")

#loop every combination
vars = colnames(north_bays[2:15])
var_pairs = combn(vars, 2) # Combinations of vars, 2 at a time

libSize = paste(10, 100, 5, collapse = " ") #have a standard library size
ccm_matrix = array(NA, dim = c(length(vars), length(vars)), dimnames = list(vars,vars)) 

for (i in 1:ncol(var_pairs)) {
ccm_out = CCM(dataFrame = north_bays, columns = var_pairs[1, i], target = var_pairs[2,i], 
              libSizes = libSize, tau= 1, E = 3,showPlot = FALSE, sample=100,replacement=TRUE) 
outVars = names(ccm_out)
var_out = unlist(strsplit(outVars[2], ":")) 
ccm_matrix[var_out[2], var_out[1]] = ccm_out[nrow(ccm_out), 2] #it should be the the coefficient at the max lib size not min!
var_out = unlist(strsplit(outVars[3], ":"))
ccm_matrix[var_out[2], var_out[1]] = ccm_out[nrow(ccm_out), 3] #it should be the the coefficient at the max lib size not min! or
}

corrplot(ccm_matrix, type="upper", method="circle", title="", mar=c(0,0,2,0), 
         #addCoef.col = 'black', 
         diag=F,tl.col = "black")
ccm_matrix

```


