---
title: "WF predators"
author: "tara"
date: "6/14/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The CTDEEP will have data going back to 1984. ---> need to request. 
The URI GSO has data to the 1960s ---> Jeremy Collie is the contact. This would probably be the best dataset to tackle this question. 
We have NMFS bottom trawl data but not idea for some inshore species. 
We have blue crab data from Tanya. 
Let's focus on CT for now because that is where our SB data is from. We might not have the searobin data yet. 
```{r}
library(dplyr)
library(kableExtra)
library(forcats)
library(Metrics)
library(tidyverse)
library(purrr)
library(rEDM)
library(corrplot)
#setwd("C:/Users/tdolan/Documents/R-Github/age_structureEDM")
```


**Load the data**
format the crab data
```{r}
BC <-read.csv('steve_17pops_summarized_landings.csv', header=TRUE)
#get blue crab data from the connectituct region. 
#Site 4 is middle of the LI sound but only goes back to 1992. 
#Site 5 is near great neck, LI 1988 ---> 2015
#Site 6 is Jamaica Bay, NY 1988 ---> 2015
#Site 3 is the Peconics  --> 1987- 2015
#Site 7 is the Hudson river near white plains --> 1988-2015
#
#Let's start with the peconics for now. 
PecBC <-filter(BC, Site==3)%>% 
 filter(Year >= 1987 & Year <= 2012)%>%
 mutate(scaled_landings = scale(Landings))
```

format the striped bass data
```{r}
#using CTSB indices for now
CTSB <- read.csv("CTtrawlabundanceage19872017.csv", header = TRUE)

#we are getting rid of the 15+ category
CTSB <-CTSB[,2:16]
pop24 <-pivot_longer(CTSB, 2:15, names_to = "age_class")%>%mutate(age_class=as.factor(age_class),lNvalue=log(value))%>%as.data.frame()%>%mutate(age_class = fct_relevel(age_class, c("X1","X2","X3","X4","X5","X6","X7","X8","X9","X10","X11","X12","X13","X14")))

#create an abundance index from all age classes
popAll <-mutate(CTSB, N_total=X1+X2+X3+X4+X5+X5+X6+X7+X8+X9+X10+X11+X12+X13+X14)
popAll <-mutate(popAll, Year=as.integer(Year))

#make a long version of popALL which you will use later. 
names(popAll)<-c("time_step","1","2","3","4","5","6","7","8","9","10","11","12","13","14","N_total")
popvert <-as.data.frame(pivot_longer(popAll, 2:15, names_to = "age",values_to="abundance"))
popvert <-popvert %>% mutate(time_step=as.numeric(time_step)) %>%
                             group_by(time_step) %>% arrange(age) %>%as.data.frame()

#find mean and SD of each column and save for later to use to unscale data
means <-colMeans(popAll[sapply(popAll, is.numeric)])
stdevs <-popAll %>% summarise_if(is.numeric, sd)
savescale <-as.data.frame(cbind(means,t(stdevs)))
names(savescale)<-c("means","sd")
savescale <-savescale[2:15,]
#scale each column
popscale <-transmute_at(popAll,2:15, scale) 
popscale <-rownames_to_column(popscale, var="time_step")

#vertical
popscv <-as.data.frame(pivot_longer(popscale, 2:15, names_to = "age",values_to="abundance"))
popscv <-popscv %>% mutate(time_step=as.numeric(time_step)) %>%
                             group_by(time_step) %>% arrange(age)
popscv <-as.data.frame(popscv)
```

format the summer flounder data
```{r}
#there are many different indices we could go with. 
#for now let's use the CTDEP fall survey and the Peconics survey
#but CTDEP fall survey is missing 2010, so maybe peconics survey? 
pecSF <-read.csv("peconicsSF.csv", header = 4TRUE)
CTSF_fall <-read.csv("SF_fall_CTDEEP.csv", header = TRUE)
CTSF_spring <-read.csv("SF_spring_CTDEEP.csv", header = TRUE)

#formating data
pecSFsc <-pecSF[,1:10]
pecSFsc <- filter(pecSFsc, Year >= 1987 & Year <= 2012)
pecSFsc <-transmute_at(pecSFsc,2:10, scale) 
pecSFsc$Year <-pecSF$Year
```

format the winter flounder data
```{r}
WFCTYOY <-read.csv("WF_CTYOY_matt.csv", header = TRUE)
#age 0 shifted forward one year making it an age 1 index (according to Matt)
WFCTYOY <-WFCTYOY[8:36,1:2]
WFCTYOY <-mutate(WFCTYOY, age0 = lead(Total,1)) %>%select(-Total)#CAREFUL HERE 

WFPec <-read.csv("WF_NYpeconicsYOY_matt.csv", header = TRUE)
WFCT_spring <-read.csv("WF_CTDEEP_Spring_matt.csv", header = TRUE)

#formatting data
WFPec <-filter(WFPec, Year >= 1987 & Year <= 2012)%>% mutate(scaled_YOY= scale(Total))
```

graph peconic blue crab vs. peconic winter flounder and peconic summer flounder (scaled).
Then try to predict YOY winter flounder from some block combination of summer flounder & blue crab. 
```{r}
Peconics <-mutate(WFPec, species="WF")%>%dplyr::select(-Total)
SFp0 <-select(pecSFsc, X0, Year) %>%rename(scaled_YOY=X0) %>%mutate(species="SF") #SF 1987-2012
PecBC0 <-select(PecBC, Year, scaled_landings)%>%rename(scaled_YOY=scaled_landings)%>%mutate(species="BC") #Blue crab 1987-2015
Peconics <-full_join(Peconics, SFp0)%>%full_join(PecBC0)%>%mutate(Year=as.numeric(as.character(Year)))
#Peconics <-filter(Peconics, Year >= 1987 & Year <= 2012) #clip the dataset. 
###You actually should probably have scaled this after clipping the dataset to between 1987 and 2012 so excluded years would not bias. 
###
Peconics%>%
 ggplot(aes(Year,scaled_YOY, color=species))+
 geom_line()+ theme_classic()

YOYwide <- pivot_wider(Peconics, id_cols=Year, names_from=species,values_from=scaled_YOY) 
YOYwide <-as.data.frame(YOYwide)

cormat=cor(YOYwide[,-1])
corrplot(cormat, method = "color", tl.cex=0.5, tl.col = "black")
```

cross mapping the peconics dataset
```{r}
#what is the embedding dimension of winter flounder
rho_E <- EmbedDimension(dataFrame = YOYwide, columns = "WF", target = "WF",lib = "1 26", pred = "1 26", showPlot = TRUE) 
# gonna go with 4. 

YOYwide <-mutate_all(YOYwide, as.numeric)
vars = colnames(YOYwide[2:4])
var_pairs = combn(vars, 2) # Combinations of vars, 2 at a time

libSize = paste(10, 100, 5, collapse = " ") #have a standard library size
ccm_matrix = array(NA, dim = c(length(vars), length(vars)), dimnames = list(vars,vars)) 

for (i in 1:ncol(var_pairs)) {
ccm_out = CCM(dataFrame = YOYwide, columns = var_pairs[1, i], target = var_pairs[2,i], 
              libSizes = libSize, tau= -1, E = 4,showPlot = TRUE, sample=100,replacement=TRUE) 
outVars = names(ccm_out)
var_out = unlist(strsplit(outVars[2], ":")) 
ccm_matrix[var_out[2], var_out[1]] = ccm_out[1, 2]
var_out = unlist(strsplit(outVars[3], ":"))
ccm_matrix[var_out[2], var_out[1]] = ccm_out[1, 3]
}

ccm_matrix
```

```{r}
ccm_matrix %>%
  round(digits=3)%>%
  kbl(caption = "CCM matrix") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```

Compare summer flounder of each age to winter flounder ages 0 and 1
```{r}
WFCT <-WFCT_spring[4:32,1:10]
WFCT <-full_join(WFCT, WFCTYOY)%>%filter(Year >=1988 & Year <= 2012)%>%select(-X) #not a lot of data here. #the age 0 is possibly from a different survey. The total may not reflect the real total now that we have combined with age 0
names(WFCT)<-c("Year","WFTotal","WFAge1","WFAge2","WFAge3","WFAge4","WFAge5","WFAge6","WFAge7plus","WFAge0")

#summer flounder spawn in the summer, so the young YOY are in the fall. 
#Young yoy of fall 1984 are the spring yoy of 1985
CTSF_fall0 <-select(CTSF_fall, Year, X0)%>%mutate(SFAge0=lag(X0,1))%>%
 filter(Year >=1988 & Year <= 2012)%>%select(-X0)
SFCT <-left_join(CTSF_fall0,CTSF_spring)%>%select(-X0)
names(SFCT)<-c("Year","SFAge0","SFAge1","SFAge2","SFAge3","SFAge4","SFAge5","SFAge6","SFAge7plus","SFTotal")

SFWF <-full_join(WFCT,SFCT)
SFWFscaled <-transmute_at(SFWF,2:19, scale) 
SFWFscaled$Year <-SFWF$Year
SFWF0 <-select(SFWFscaled, -c("WFAge3","WFAge4","WFAge5","WFAge6","WFAge7plus"))

SFWFlong <-pivot_longer(SFWFscaled, cols=c("WFTotal",  "WFAge1", "WFAge2", "WFAge3", "WFAge4",     "WFAge5",  "WFAge6","WFAge7plus","WFAge0", "SFAge0","SFAge1", "SFAge2",  "SFAge3",     "SFAge4",     "SFAge5","SFAge6", "SFAge7plus", "SFTotal"), names_to = "Age",values_to = "abundance")

SFWFlong%>%
 ggplot(aes(Year,abundance))+
 geom_line()+ facet_wrap(~Age)+
 theme_classic()

SFWF0 <-mutate_all(SFWF0, as.numeric)
SFWF0 <-na.omit(SFWF0)
cormat=cor(SFWF0[,1:13])
corrplot(cormat, method = "color", tl.cex=0.5, tl.col = "black")
```

cross mapping the SF and young WF dataset
```{r}
#what is the embedding dimension of winter flounder? We will base it on WF total?
SFWF0 <-SFWF0[,c(14, 1,2,3,4,5,6,7,8,9,10,11,12,13)]
rho_E <- EmbedDimension(dataFrame = SFWF0, columns = "WFTotal", target = "WFTotal",lib = "1 24", pred = "1 24", showPlot = TRUE) 
# gonna go with 4. 

#YOY
rho_E <- EmbedDimension(dataFrame = SFWF0, columns = "WFAge0", target = "WFAge0",lib = "1 24", pred = "1 24", showPlot = TRUE) 
#E = 3

#Age1
rho_E <- EmbedDimension(dataFrame = SFWF0, columns = "WFAge1", target = "WFAge1",lib = "1 24", pred = "1 24", showPlot = TRUE)
#E=4

#Age2
rho_E <- EmbedDimension(dataFrame = SFWF0, columns = "WFAge2", target = "WFAge2",lib = "1 24", pred = "1 24", showPlot = TRUE)
#E=4, but could be E=1
```

Next time: Feed each one it's own embedding dimension
```{r}
vars = colnames(SFWF0[2:13])
var_pairs = combn(vars, 2) # Combinations of vars, 2 at a time

libSize = paste(10, 100, 5, collapse = " ") #have a standard library size
ccm_matrix = array(NA, dim = c(length(vars), length(vars)), dimnames = list(vars,vars)) 

for (i in 1:ncol(var_pairs)) {
ccm_out = CCM(dataFrame = SFWF0, columns = var_pairs[1, i], target = var_pairs[2,i], 
              libSizes = libSize, tau= -1, E = 4,showPlot = TRUE, sample=100,replacement=TRUE) 
outVars = names(ccm_out)
var_out = unlist(strsplit(outVars[2], ":")) 
ccm_matrix[var_out[2], var_out[1]] = ccm_out[1, 2]
var_out = unlist(strsplit(outVars[3], ":"))
ccm_matrix[var_out[2], var_out[1]] = ccm_out[1, 3]
}

ccm_matrix

```

```{r}
ccm_matrix %>%
  round(digits=3)%>%
  kbl(caption = "CCM matrix") %>%
  kable_classic(full_width = F, html_font = "Cambria")
```   

Compare SB and WF in CT spring trawl
```{r}
WFCT <-WFCT_spring[4:32,1:10]
WFCT <-full_join(WFCT, WFCTYOY)%>%select(-X) #not a lot of data here. #the age 0 is possibly from a different survey. The total may not reflect the real total now that we have combined with age 0
names(WFCT)<-c("Year","WFTotal","WFAge1","WFAge2","WFAge3","WFAge4","WFAge5","WFAge6","WFAge7plus","WFAge0") #1984- 2016
WFCT <-filter(WFCT, Year >= 1987 & Year <= 2012)

#now the striped bass
SBCT <-popscale
names(SBCT) <-c("Year", "SBAge1","SBAge2","SBAge3","SBAge4","SBAge5","SBAge6","SBAge7","SBAge8","SBAge9","SBAge10","SBAge11","SBAge12","SBAge13","SBAge14")
SBCT <-SBCT[,-1]
SBCT$Year <-CTSB$Year  #1987-2017
SBCT <-filter(SBCT, Year >= 1987 & Year <= 2012)

CTWFSB <-full_join(SBCT,WFCT) %>%
 mutate(SBTotal = SBAge1+SBAge2+SBAge3+SBAge4+SBAge5+SBAge6+SBAge7+SBAge8+SBAge9+SBAge10+SBAge11+SBAge12+SBAge13+SBAge14)%>% select(-Year)%>%mutate_all(scale)
CTWFSB$Year <-SBCT$Year

CTWFSB0 <-select(CTWFSB, -c("WFAge3","WFAge4","WFAge5","WFAge6","WFAge7plus"))

CTSBlong <-pivot_longer(CTWFSB, cols=c("WFTotal",  "WFAge1", "WFAge2", "WFAge3", "WFAge4",     "WFAge5",  "WFAge6","WFAge7plus","WFAge0","SBAge1","SBAge2","SBAge3","SBAge4","SBAge5","SBAge6","SBAge7","SBAge8","SBAge9","SBAge10","SBAge11","SBAge12","SBAge13","SBAge14", "SBTotal"), names_to = "Age",values_to = "abundance")

CTSBlong%>%
 ggplot(aes(Year,abundance))+
 geom_line()+ facet_wrap(~Age)+
 theme_classic()

CTWFSB0 <-mutate_all(CTWFSB0, as.numeric)
CTWFSB0c <-na.omit(CTWFSB0)%>% select(-Year)
cormat=cor(CTWFSB0c)
corrplot(cormat, method = "color", tl.cex=0.5, tl.col = "black")

```

Next time: Feed each one it's own embedding dimension
```{r}
CTWFSB0 <-CTWFSB0[,c(20,1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19)]
vars = colnames(CTWFSB0[2:20])
var_pairs = combn(vars, 2) # Combinations of vars, 2 at a time

libSize = paste(10, 100, 5, collapse = " ") #have a standard library size
ccm_matrix = array(NA, dim = c(length(vars), length(vars)), dimnames = list(vars,vars)) 

for (i in 1:ncol(var_pairs)) {
ccm_out = CCM(dataFrame = CTWFSB0, columns = var_pairs[1, i], target = var_pairs[2,i], 
              libSizes = libSize, tau= -1, E = 4,showPlot = TRUE, sample=100,replacement=TRUE) 
outVars = names(ccm_out)
var_out = unlist(strsplit(outVars[2], ":")) 
ccm_matrix[var_out[2], var_out[1]] = ccm_out[1, 2]
var_out = unlist(strsplit(outVars[3], ":"))
ccm_matrix[var_out[2], var_out[1]] = ccm_out[1, 3]
}

ccm_matrix

```

CT lists totals by species. 
```{r}
WFSBtots <-select(CTWFSB0, Year, WFAge0, WFTotal, SBTotal)
SFWF2 <-select(SFWF0, Year, SFTotal)

WFSBtots <-full_join(WFSBtots,SFWF2)
WFSBtots <-WFSBtots[2:24,]

Wftotslong <-pivot_longer(WFSBtots, cols=c("WFAge0",  "WFTotal", "SBTotal", "SFTotal"), names_to = "Age",values_to = "abundance")

Wftotslong%>%
 ggplot(aes(Year,abundance))+
 geom_line()+ facet_wrap(~Age)+
 theme_classic()
```

```{r}
vars = colnames(WFSBtots[2:5])
var_pairs = combn(vars, 2) # Combinations of vars, 2 at a time

libSize = paste(10, 100, 5, collapse = " ") #have a standard library size
ccm_matrix = array(NA, dim = c(length(vars), length(vars)), dimnames = list(vars,vars)) 

for (i in 1:ncol(var_pairs)) {
ccm_out = CCM(dataFrame = WFSBtots, columns = var_pairs[1, i], target = var_pairs[2,i], 
              libSizes = libSize, tau= -1, E = 4,showPlot = TRUE, sample=100,replacement=TRUE) 
outVars = names(ccm_out)
var_out = unlist(strsplit(outVars[2], ":")) 
ccm_matrix[var_out[2], var_out[1]] = ccm_out[1, 2]
var_out = unlist(strsplit(outVars[3], ":"))
ccm_matrix[var_out[2], var_out[1]] = ccm_out[1, 3]
}

ccm_matrix
```

